

# Set up

Import packages
```{r}
suppressPackageStartupMessages({
  suppressWarnings({
    library(tidyverse)
    library(readxl)
    library(here)
    library(sf)
  })})
```

Set settings and universal variables
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here())

options(scipen = 999)
```

# Import data


## Crosswalk

These are the drinking water demand units in CalSim we want to focus on
```{r}
crosswalk = read.csv("Outputs/Crosswalk demand and delivery IDs/20260413_crosswalk_processed.csv")
```

## Network data

Import most recent version and make a couple manual edits:
- There are 12 ag-demand units that aren't relevant to our drinking water calculations, so we'll change them to 'Out of CalSim'
- There is also an infinite loop between CA4410014 and CA4410002 since they both sell water to each other. Kristin confirmed that none of those connections relate to CalSim, so we'll also change those connections to 'Out of CalSim' to avoid the infinite loop
```{r}
filter_out = c("20_NA1", "25_PA1; SIDSH", "60N_NA1", "61_NA2", "64_PA1", "73_PA1", "73_PA3", "90_PA1", "90_PA2", "CLLPT", "FKC")

network_df = read.csv("C:/Users/benji/Desktop/COEQWAL/ca-surface-water-sources/Data/Outputs/20260505_network_joined.csv")%>%
  mutate(average_source_usage = average_source_usage / 100,
         calsim_demandunit = str_squish(calsim_demandunit),
         calsim_demandunit = ifelse(
           (pwsid %in% c("CA4410014", "CA4410002") & calsim_demandunit == "PWSID") | calsim_demandunit %in% filter_out,
           "Out of CalSim", calsim_demandunit))
```


# Process data

Create new rows representing GW deliveries to fill in the remainder (to hit 100% volume) for each system
```{r}
# Combine with original data
network_df = bind_rows(
  network_df, 
  network_df %>%
    group_by(pwsid) %>%
    reframe(
      source_name = "gw", 
      source_type = "gw",
      calsim_demandunit = "gw",
      average_source_usage = 1 - sum(average_source_usage))%>%
    filter(average_source_usage > 0))
```

# Demand unit tracing

Create a dataframe of 'end nodes' that have a specified demand unit (and therefore don't need to be traced back further)
```{r}
end_nodes <- network_df %>%
  filter(calsim_demandunit != "PWSID") %>%
  select(pwsid, traced_du = calsim_demandunit, average_source_usage)%>%
  group_by(pwsid, traced_du)%>%
  reframe(average_source_usage = sum(average_source_usage))
```

Recursive function to trace back when any connections have their demand unit listed as 'PWSID'
```{r}
# Step 1: Build PWSID → DU mappings (terminal nodes)
du_mappings <- network_df %>%
  filter(calsim_demandunit != "PWSID") %>%
  select(pwsid, source = calsim_demandunit, volume = average_source_usage)

# Step 2: Build PWSID → PWSID mappings (intermediate nodes)
pwsid_mappings <- network_df %>%
  filter(calsim_demandunit == "PWSID") %>%
  select(pwsid, source = source_name, volume = average_source_usage)

# Step 3: Combine and calculate fractions of volume
complete_map <- bind_rows(du_mappings, pwsid_mappings) %>%
  group_by(pwsid, source) %>%
  summarise(volume = sum(volume), .groups = "drop") %>%
  group_by(pwsid) %>%
  mutate(fraction = volume / sum(volume)) %>%
  ungroup()

# Step 4: Create lookup list
complete_list <- complete_map %>%
  group_by(pwsid) %>%
  group_split() %>%
  setNames(map_chr(., ~ first(.x$pwsid))) %>%
  map(~ select(.x, source, fraction))

# Step 5: Recursive trace function
trace_to_dus <- function(source_name, volume = 1, complete_list, visited = c()) {
  # Prevent infinite loops from circular dependencies
  if (source_name %in% visited) {
    warning(paste("Circular reference detected:", source_name))
    return(tibble(traced_du = source_name, adjusted_volume = volume))}
  
  # If this source is a PWSID that can be decomposed
  if (source_name %in% names(complete_list)) {
    components <- complete_list[[source_name]]
    
    results <- map2_dfr(
      components$source,
      components$fraction,
      ~ trace_to_dus(.x, volume * .y, complete_list, c(visited, source_name)))
    return(results)} 
  else {
    # End node, so stop recursion and record that endnode's info
    return(tibble(traced_du = source_name, adjusted_volume = volume))}}

# Step 6: Apply to all purchases
purchases <- network_df %>%
  filter(calsim_demandunit == "PWSID") %>%
  select(pwsid, source_name, average_source_usage)

traced_results <- purchases %>%
  rowwise() %>%
  mutate(
    traced = list(trace_to_dus(source_name, average_source_usage, complete_list))) %>%
  unnest(traced) %>%
  select(pwsid, traced_du, adjusted_volume) %>%
  group_by(pwsid, traced_du) %>%
  summarise(average_source_usage = sum(adjusted_volume), .groups = "drop")

# Step 7: combine traced results with end nodes to get 'master sheet' of each system's complete volume distribution across demand units
final = rbind(traced_results, end_nodes)%>%
  mutate(average_source_usage = average_source_usage *100)%>%
  group_by(pwsid, traced_du)%>%
  reframe(average_source_usage = sum(average_source_usage))
```

## Diagnostics

Do some manual tracing to make sure everything was calculated correctly
```{r, eval = F}
final%>%
  filter(pwsid == "CA0110003")
```

```{r, eval = F}
network_df%>%
  mutate(average_source_usage = average_source_usage * 100)%>%
  filter(pwsid %in% c("CA0110003", "CA0110010"))%>%
  select(pwsid, source_name, average_source_usage, calsim_demandunit)
```

## Filter out GW and 'Out of CalSim' demand units
```{r}
traced_dus = final%>%
  filter(!traced_du %in% c("gw", "Out of CalSim", "Out of Calsim"))
```



## Separate combined demand units

In many cases, multiple demand units will be listed, separate by a semicolon. In these cases, we want to split them up and distribute the volume between them proportional to their historical deliveries. 

## Calculate historical deliveries

So first, let's import the annual delivery data from CalSim (baseline scenario s0020) and calculate the mean annual delivery per demand unit. 

```{r}
extract_scenario_id = function(filename) {
  str_extract(filename, "s\\d+") %>% tolower()}

file_path = "Inputs/Deliveries/s0020_DCRadjBL_2020LU_wTUCP_DELIVERIES.csv"

scenario_id = extract_scenario_id(basename(file_path))
  
data = read.csv(file_path) %>%
    mutate(row_id = row_number())%>%
    pivot_longer(cols = -row_id,
                 names_to = "col_position",
                 values_to = "value")%>%
    pivot_wider(
      names_from = row_id,
      values_from = value)%>%
    rename(
      col_position = 1,
      delivery_id = 2,    
      units = 7)
  
dates = data%>%
    slice(1) %>%  
    select(8:ncol(.)) %>%
    pivot_longer(
      cols = everything(),
      names_to = "date_id",
      values_to = "date")
  
historical_deliveries = data%>%
    filter(units == "TAF" & delivery_id %in% crosswalk$delivery_id)%>%
    select(-c(1,3:units))%>%
    pivot_longer(
      cols = -delivery_id,
      names_to = "date_id",
      values_to = "delivery_taf")%>%
    left_join(dates, by = "date_id")%>%
    mutate(scenario_id = scenario_id,
           delivery_taf = as.numeric(ifelse(delivery_taf == "", NA, delivery_taf)),
           year = year(date))%>%
    select(scenario_id, delivery_id, year, delivery_taf) %>%
    left_join(crosswalk%>%select(dwuc, delivery_id), by = "delivery_id")%>%
    group_by(dwuc, year)%>%
    reframe(annual_delivery_taf = sum(delivery_taf))%>%
    group_by(dwuc)%>%
    reframe(mean_annual_delivery_taf = mean(annual_delivery_taf))
```

Now, we can separate out any rows with semicolons in the demand unit. 
```{r}
traced_dus_split = traced_dus%>%
  mutate(traced_du = str_replace_all(traced_du, "[\n]+", ""),
         split = ifelse(grepl(";", traced_du), 1, 0))%>%
  separate_rows(traced_du, sep = ";\\s*")%>%
  left_join(historical_deliveries, by = c("traced_du" = "dwuc"))%>%
  group_by(pwsid) %>%
  mutate(total_deliveries = sum(mean_annual_delivery_taf))%>%
  ungroup()%>%
  mutate(ratio = mean_annual_delivery_taf / total_deliveries,
    average_source_usage = ifelse(
      split == 1 & !is.na(mean_annual_delivery_taf), average_source_usage*ratio, average_source_usage))
```


## Export
```{r}
write.csv(traced_dus_split, "Outputs/Static demand/20260505_traced_demand_units.csv", row.names = F)
```
