Refer to [project overview](https://docs.google.com/document/d/1QPh7cKy7fmokalWuou5Z_zqG6680rybq8CC6gmiC8RE/edit?tab=t.0) document for guidance


# Set up

Import packages
```{r}
suppressPackageStartupMessages({
  suppressWarnings({
  library(tidyverse)   # Data wrangling
  library(here)        # File management
  library(readxl)      # Read excel files
  })})
```

Set settings and universal variables
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here())
options(scipen = 999)
```

# Import and process data

## Crosswalk
Crosswalk between demand units' demand ID and delivery ID
Provided by Kristin via [this folder](https://drive.google.com/drive/folders/1BYFQlIA0JJ31sAHDt7yMovLWWZvwoMeO)
```{r}
crosswalk = read_excel("Inputs/Master crosswalk SW DUs M&I.xlsx")%>%
  rename(dwuc = 1, demand_id = 2, delivery_id = 3)
```

Demand and delivery data downloaded from the respective Google Drive folder linked to each scenario in [this spreadsheet](https://docs.google.com/spreadsheets/d/1iXqEc4sMVC8gKqPzhf_VOS32EXKzocAP2pNJ5PZlFbY/edit?gid=0#gid=0). Once in a scenario's folder, click on 'Data Extraction', then 'Demands_Deliveries'. We want to download csvs of monthly values for both demand and deliveries.


## Monthly demand data

Need to standardise units and join to crosswalk to get demand unit ID
```{r}
raw = read.csv("Inputs/Demands/s0045_DCRadjBL_2020LU_RelaxFallX2_DEMANDS.csv")

processed_wide = raw%>%
  mutate(row_id = row_number())%>%
  pivot_longer(cols = -row_id,
               names_to = "col_position",
               values_to = "value")%>%
  pivot_wider(
    names_from = row_id,
    values_from = value)%>%
  rename(
    col_position = 1,
    demand_id = 2,    
    units = 7)

dates = processed_wide %>%
  slice(1) %>%
  select(8:ncol(.)) %>%
  pivot_longer(
    cols = everything(),
    names_to = "date_id",
    values_to = "date")

demand_processed = processed_wide%>%
  filter(units == "TAF" & demand_id %in% crosswalk$demand_id)%>%
  select(-c(1,3:units))%>%
  pivot_longer(
    cols = -demand_id,
    names_to = "date_id",
    values_to = "demand_taf")%>%
  mutate(demand_taf = as.numeric(ifelse(demand_taf == "", 0, demand_taf)))%>%
  left_join(., dates, by = "date_id")%>%
  select(-date_id)%>%
  left_join(., crosswalk %>% select(-delivery_id), by = "demand_id")

demand_processed%>%
  group_by(demand_id, dwuc)%>%
  reframe(sum = sum(demand_taf))%>%
  arrange(sum)
```


## Deliveries

Need to standardise units and join to crosswalk to get demand unit ID
```{r}
raw = read.csv("Inputs/Deliveries/s0045_DCRadjBL_2020LU_RelaxFallX2_DELIVERIES.csv")

processed_wide = raw%>%
  mutate(row_id = row_number())%>%
  pivot_longer(cols = -row_id,
               names_to = "col_position",
               values_to = "value")%>%
  pivot_wider(
    names_from = row_id,
    values_from = value)%>%
  rename(
    col_position = 1,
    delivery_id = 2,    
    units = 7)

dates = processed_wide %>%
  slice(1) %>%  # Get the row with dates
  select(8:ncol(.)) %>%
  pivot_longer(
    cols = everything(),
    names_to = "date_id",
    values_to = "date")

delivery_processed = processed_wide%>%
  filter(units == "TAF" & delivery_id %in% crosswalk$delivery_id)%>%
  select(-c(1,3:units))%>%
  pivot_longer(
    cols = -delivery_id,
    names_to = "date_id",
    values_to = "delivery_taf")%>%
  mutate(delivery_taf = as.numeric(ifelse(delivery_taf == "", 0, delivery_taf)))%>%
  left_join(., dates, by = "date_id")%>%
  left_join(crosswalk %>% select(-demand_id), by = "delivery_id")
```

```{r}
combined = full_join(delivery_processed, demand_processed, by = c("dwuc", "date"))%>%
  mutate(percent_demand_met = ifelse(delivery_taf >= demand_taf, 100, delivery_taf / demand_taf * 100),
         year = year(date))%>%
  select(dwuc, delivery_id, demand_id, year, date, delivery_taf, demand_taf, percent_demand_met)

combined%>%
  filter(delivery_taf > 0 & demand_taf == 0)
```

# Analysis

## Missingness

Create list of demand units that are entirely missing either delivery or demand values.
```{r}
missing = crosswalk%>%
  mutate(demand_missing = ifelse(!demand_id %in% demand_processed$demand_id, 1, 0),
         delivery_missing = ifelse(!delivery_id %in% delivery_processed$delivery_id, 1, 0))%>%
  filter(delivery_missing == 1 | demand_missing == 1)
```

Investigate month to month missingness for demand & delivery values
```{r}
combined%>%
  mutate(delivery_id_missing = ifelse(is.na(delivery_id), 1, 0),
         delivery_taf_missing = ifelse(is.na(delivery_taf), 1, 0),
         demand_id_missing = ifelse(is.na(demand_id), 1, 0),
         demand_taf_missing = ifelse(is.na(demand_id), 1, 0))%>%
  group_by(dwuc, year)%>%
  reframe(delivery_id_missing = sum(delivery_id_missing),
          delivery_taf_missing = sum(delivery_taf_missing),
          demand_id_missing = sum(demand_id_missing),
          demand_taf_missing = sum(demand_taf_missing))%>%
  filter(delivery_taf_missing > 0 | demand_taf_missing > 0)%>%
  filter(!dwuc %in% missing$dwuc)

combined%>%
  filter(!dwuc %in% missing$dwuc)%>%
  mutate(missing = ifelse(is.na(delivery_taf) | is.na(demand_taf), 1, 0))%>%
  group_by(year, dwuc)%>%
  reframe(annual_delivery = sum(delivery_taf),
          annual_demand = sum(demand_taf),
          n_missing_months = sum(missing))%>%
  ungroup()%>%
  mutate(totally_missing = ifelse(n_missing_months == 12, 1, 0),
         partially_missing = ifelse(n_missing_months > 0, 1, 0))%>%
  group_by(dwuc)%>%
  reframe(years_totally_missing = sum(totally_missing),
          years_partially_missing = sum(partially_missing))%>%
  arrange(-years_totally_missing, -years_partially_missing)
```

Exclude any demand units that are entirely missing demand or delivery values. That way, we avoid scenarios where demand has a value and delivery is missing, which could incorrectly throw a 0% met when that's not necessarily true. 
Similarly, exclude any months for demand units where either demand or delivery is NA.
```{r}
combined = combined%>%
  filter(!dwuc %in% missing$dwuc)
  mutate(missing = ifelse(is.na(delivery_taf) | is.na(demand_taf), 1, 0))%>%
  filter(missing == 0)
  
n_distinct(combined$dwuc)
```


## Percent demand met and tiers

Aggregate delivery and demand by year, then calculate % of demand met by year. 

Based on that, calculate % of years where demand is fully met and the minimum % of demand met in deficit years.
From there, we can assign tiers according to the criteria we've listed in [this spreadsheet](https://docs.google.com/spreadsheets/d/1xcQIR_J96-cs7BuCrXjznwkinLgxl-Pf9tA3mJ2GiyA/edit?gid=916228682#gid=916228682)
```{r}
combined_by_year = combined%>%
  group_by(year, dwuc)%>%
  reframe(annual_delivery = sum(delivery_taf),
          annual_demand = sum(demand_taf))%>%
  ungroup()%>%
  mutate(percent_demand_met = ifelse(annual_delivery >= annual_demand, 100, annual_delivery / annual_demand * 100))

#write.csv(combined_by_year, "Outputs/insert scenario id here_demand_met_by_year.csv", row.names = F)

tiers = combined_by_year%>%
  mutate(demand_met = ifelse(percent_demand_met >= 100, 1, 0))%>%
  group_by(dwuc)%>%
  reframe(perc_years_met = sum(demand_met)/n()*100,
          min_met = min(percent_demand_met))%>%
  ungroup()%>%
    mutate(tier = case_when(
    perc_years_met >= 90 & min_met > 75 ~ 1,
    perc_years_met >= 70 & min_met > 75 ~ 2,
    perc_years_met >= 50 & min_met > 50 ~ 3,
    T ~ 4))

# rather than separate csv of tiers for each scenario, combine all scenarios so that we have one row per scenario and one column per demand unit (dwuc). and then the value in the corresponding cell is the tier (1, 2, 3, 4)
```


```{r}
read.csv("Outputs/Annual demand met/s0010_demand_met_by_year.csv")%>%
  filter(dwuc == "02_PU")

read.csv("Outputs/Monthly demand met/s0010_demand_met_by_month.csv")%>%
  filter(dwuc == "02_PU" & year < 1923)
  summarise(sum = sum(delivery_taf))

read.csv("Inputs/Deliveries/s0010_draftDCP_NDD_SWP6000_Beth_2020_DELIVERIES.csv", skip = 1)%>%
  select(contains("DN_02_PU"))
  filter(dwuc == "02_PU" & year == 1921)%>%
  summarise(sum = sum(delivery_taf))
```

```{r}
read.csv("C:/Users/benji/Downloads/s0010_draftDCP_NDD_SWP6000_Beth_2020_DELIVERIES-ANNUAL.csv", skip = 1)%>%
  select(DN_02_PU)
```


Missingness
```{r}
read.csv("Outputs/Missing demand units/s0010_missing_dus.csv")
  reframe(dem = sum(demand_missing), del = sum(delivery_missing))

read.csv("Inputs/Deliveries/s0010_draftDCP_NDD_SWP6000_Beth_2020_DELIVERIES.csv", skip = 1)%>%
  select(contains("D_CLRLK_CLLPT"))
read.csv("Inputs/Demands/s0010_draftDCP_NDD_SWP6000_Beth_2020_DEMANDS.csv", skip = 1)%>%
  select(contains("U_CLPT"))
```

