

# Set up

Import packages
```{r}
suppressPackageStartupMessages({
  suppressWarnings({
    library(tidyverse)
    library(readxl)
    library(here)
    library(sf)
  })})
```

Set settings and universal variables
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here())

options(scipen = 999)
```

# Import data

## Crosswalk
These are the drinking water demand units in CalSim we want to focus on
```{r}
crosswalk = read.csv("Outputs/Crosswalk demand and delivery IDs/20260409_crosswalk_processed.csv")
```

## Traced demand units
Dataset generated in 'system_demand_unit_tracing_loop.Rmd' in which we traced all systems back to all demand units they are serviced by
```{r}
traced_dus_split = read.csv("Outputs/Static demand/20260410_traced_demand_units.csv")
```


# Calculating demand by demand unit

Great, so now, for each system, we know the distribution of how much water they get from different demand units. 
So now, in order to calculate total demand by demand unit, we can join in a dataset that we calculated of demand (in TAF) per year per system. 

I am drawing upon output genereated by the 'calculating_demand_by_system.Rmd' script, which is based on Jenny's original code and methodology
that can be found [here](https://github.com/jenrempel/coeqwal_demand/tree/main). 

```{r}
demand_by_system = read.csv("Outputs/Static demand/20260409_static_demand_by_system.csv")%>%
  select(pwsid, demand_annual_taf)%>%
  rename(demand_system_af = 2)
```

```{r}
demand_by_du = left_join(traced_dus_split, demand_by_system, by = "pwsid")%>%
  mutate(demand_du =  demand_system_af * average_source_usage/100)%>%
  group_by(traced_du)%>%
  reframe(demand_du_taf = sum(demand_du))%>%
  arrange(traced_du)
```

## Diagnostics
Ensure we have estimates for all demand units
```{r}
crosswalk%>%filter(!dwuc %in% demand_by_du$traced_du)
crosswalk%>%filter(!dwuc %in% traced_dus_split$traced_du)
```

## Export
```{r}
write.csv(demand_by_du, "Outputs/Static demand/20260410_static_demand_by_du_updated.csv", row.names = F)
```
