
Please refer to [this google doc](https://docs.google.com/document/d/1GPkFW5r0EbUXbv3PATY_WgLfc7k7e8fudJi01nbmGDo/edit?tab=t.0#heading=h.pgyusvq7cjo3), which is the original write-up of methods for Jenny's process in 2025, which I used as a reference.

# Set up

Import packages
```{r}
suppressPackageStartupMessages({
  suppressWarnings({
    library(tidyverse)
    library(readxl)
    library(here)
    library(sf)
  })})
```

Set settings and universal variables
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here())

options(scipen = 999)
```

# Import data

We'll import 2 datasets:
1) a list of all systems we need demands calculated for
2) our network dataset (which contains useful information such as retail/wholesale status)
```{r}
all_systems = read.csv("Outputs/Static demand/20260409_traced_demand_units.csv")%>%
  select(pwsid)%>%
  unique()

network_df = read.csv("C:/Users/benji/Desktop/COEQWAL/coeqwal-benji/Data/Outputs/20260330_network_joined.csv")

sdwis = read_excel("Inputs/SABL/EPASDWIS_Q12025_allactiveCAPWS.xlsx", skip = 3)%>%
  rename(pwsid = 1, pop_served_sdwis = 6)%>%
  select(1, 6)
```


# Join system boundaries to hydrologic region

First, let's import a shapefile of polygons representing California's hydrological regions, downloaded from [this page](https://data.ca.gov/dataset/i03-hydrologic-regions)
```{r}
hydro_regions <- read_sf("Inputs/Hydrological regions/dwr_hydro_regions/i03_Hydrologic_Regions/i03_Hydrologic_Regions.shp")%>%
  st_transform(crs = "EPSG:3857")%>%
  select(HR_NAME)%>%
  rename(hydro_region = 1)
```

We also import a shapefile of drinking water system area boundaries, obtained from [the California Stae Geoportal](https://gis.data.ca.gov/datasets/waterboards::california-drinking-water-system-area-boundaries/about).

We filter for only community systems, and since some systems have both jurisdictional and water service area boundaries, we choose just one polygon per system to avoid duplicates (choosing water service area boundary if available, and jurisdictional boundary if not).

Finally, we perform a spatial join to assign each system to the hydrological region it has the greatest area of overlap with.
```{r}
sabl = read_sf("Inputs/SABL/California_Drinking_Water_System_Area_Boundaries/California_Drinking_Water_System_Area_Boundaries.shp")%>%
  rename(pwsid = 2)%>%
  filter(STATE_CLAS == "COMMUNITY")%>%
  group_by(pwsid)%>%
  slice_max(BOUNDARY_T == "Water Service Area", n = 1)%>%
  select(pwsid)%>%
  st_transform(crs = "EPSG:3857")%>%
  st_join(hydro_regions, largest = TRUE)
```

# Calculate demand from Clearinghouse data

## Import and process
Let's obtain demand from clearinghouse where possible. 
First, we'll import the dataset, which was downloaded from [this drive folder](https://drive.google.com/drive/u/0/folders/1Rma7KnAtB27FufcOi3jIxTPj8FabI5hp).
We can then filter for only community water systems (which we did for the SABL dataset already), and clean up some of the NA values.
```{r}
clearinghouse <- readRDS("jenny_demand_by_system/processed_data/ch_wsi_2025.rds")%>%
  select(pwsid, potable_demand_unit_of_measure, total_annual_potable_demand, population, reporting_period_end_date)%>%
  filter(pwsid %in% sabl$pwsid)%>%
  mutate(across(c(potable_demand_unit_of_measure, total_annual_potable_demand), ~na_if(., "Not Available")),
         total_annual_potable_demand = as.numeric(total_annual_potable_demand))
```

## Calculate gallons per person per day
Next, we'll convert all units for demand to gallons. For any units listed as NA, we'll assume it's the same as the rest of the units for that system (if all the units for other rows in that system are the same).

Then, we'll convert all values to demand per person per day by dividing by 365 (# of days per year) and the system's population served.

We then filter for only values of gallons/per person/per day between 50-250, assuming that any values outside this range are implausible and likely due to data entry errors for the demand or unit variables. 

Finally, since there are multiple observations for many systems, we take the median across all a system's values. 
```{r}
clearinghouse_median_gpcd = clearinghouse%>%
  mutate(
    total_annual_potable_demand_in_g = case_when(
      potable_demand_unit_of_measure == "Gallons (G)" ~ total_annual_potable_demand,
      potable_demand_unit_of_measure == "Acre Feet (AF)" ~ total_annual_potable_demand * 325851,
      potable_demand_unit_of_measure == "Hundred Cubic Feet (HCF)" ~ total_annual_potable_demand * 748.05,
      potable_demand_unit_of_measure == "Million Gallons (MG)" ~ total_annual_potable_demand * 1000000,
      T ~ NA))%>%
  group_by(pwsid) %>%
  mutate(
    potable_demand_unit_of_measure = if_else(
      is.na(potable_demand_unit_of_measure) & n_distinct(potable_demand_unit_of_measure[!is.na(potable_demand_unit_of_measure)]) == 1,
      first(potable_demand_unit_of_measure[!is.na(potable_demand_unit_of_measure)]),
      potable_demand_unit_of_measure)) %>%
  ungroup()%>%
  mutate(total_annual_potable_demand_in_g_per_cap_day = total_annual_potable_demand_in_g / population / 365)%>%
  filter(total_annual_potable_demand_in_g_per_cap_day > 50,
         total_annual_potable_demand_in_g_per_cap_day < 250)%>%
  group_by(pwsid)%>% 
  reframe(median_gpcd = median(total_annual_potable_demand_in_g_per_cap_day, na.rm = T))
```


# Combine all data

Join our list of systems with the empirical demand obtained from Clearinghouse data.

Also join in our SABL layer with hydrological region info, and fill in the hydrological region manually for 5 systems that aren't in the SABL database. These were manually looked up by Jenny. 

We calculate the median demand for each hydrological region, and for any systems without empirical demand data from Clearinghouse, we fill in the missing values with the median demand for the hydrological region it's in. 

Finally, we join in the network data in order to see the provider type (Retail/Wholesale) for each system, and if systems with missing demand only do wholesale, we fill in their demand as zero (since they don't actually serve any retail population)

```{r}
combined = full_join(all_systems, clearinghouse_median_gpcd, by = "pwsid")%>%
  left_join(., sabl%>%st_drop_geometry(), by = "pwsid")%>%
  mutate(hydro_region = case_when(
    pwsid == "CA0110010" ~ "San Francisco Bay", # livermore - SF Bay
    pwsid == "CA2810008" ~  "San Francisco Bay", # yountville - SF Bay
    pwsid == "CA3010071" ~  "South Coast", # laguna beach - south coast
    pwsid == "CA3510007" ~  "Central Coast", # hollister - central coast
    pwsid == "CA3610019" ~  "South Coast", # san bernardino - south coast
    pwsid == "CA3710707" ~ "South Coast", #San clemente island --> south coast
    T ~ hydro_region))%>%
  group_by(hydro_region)%>%
  mutate(regional_median = ifelse(is.na(hydro_region), NA, median(median_gpcd, na.rm = T)))%>%
  ungroup()%>%
  filter(pwsid %in% all_systems$pwsid)%>%
  left_join(., network_df%>%select(pwsid, system_name, provider_type)%>%unique(), by = "pwsid")%>%
  left_join(., clearinghouse%>%select(pwsid, population)%>%unique(), by = "pwsid")%>%
  left_join(., sdwis, by = "pwsid")%>%
  mutate(
    approach = case_when(
      is.na(median_gpcd) & !is.na(hydro_region) ~ "imputed",
      !is.na(median_gpcd) ~ "empirical",
      provider_type == "Wholesale" ~ "wholesaler",
      T ~ NA),
    median_gpcd = case_when(
      is.na(median_gpcd) & provider_type == "Wholesale" ~ 0,
      is.na(median_gpcd) ~ regional_median, 
      T ~ median_gpcd),
    pop_source = ifelse(is.na(population), "sdwis", "clearinghouse"),
    population = ifelse(is.na(population), pop_served_sdwis, population),
    demand_annual_taf = case_when(
      approach == "wholesaler" ~ 0,
      T ~ median_gpcd*population*365/325851/1000))%>%
  select(pwsid, system_name, provider_type, hydro_region, population, pop_source, median_gpcd, approach, demand_annual_taf)
```

# Export
```{r}
write.csv(combined, "Outputs/Static demand/20260409_static_demand_by_system.csv", row.names = F)
```


# Compare with Jenny's older values
```{r}
jenny = read.csv("Inputs/Static demands/jenny_du_demand_final_123025.csv")%>%
  select(pwsid, annual_total_demand_in_gpcd_final)%>%
  unique()

comparison = combined%>%
  select(pwsid, median_gpcd, approach)%>%
  left_join(., jenny, by = "pwsid")

comparison

write.csv(comparison, "Outputs/Static demand/compare_jenny_vs_new_system_demands.csv", row.names = F)
```


