

# Set up

Import packages
```{r}
suppressPackageStartupMessages({
  suppressWarnings({
    library(tidyverse)
    library(readxl)
    library(here)
    library(sf)
  })})
```

Set settings and universal variables
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here())

wgs_crs = 4326
ca_crs = 3310
```

# Import and process data

Load in list of demand units relevant to surface water in CalSim
```{r}
calsim_dus = read_excel("Inputs/Master crosswalk SW DUs M&I.xlsx")%>%
  rename(demand_unit = 1)%>%
  select(1)%>%
  mutate(demand_unit = str_squish(demand_unit))%>%
  filter(!demand_unit %in% c("71_NU", "72_NU"))
```

Surface water demand unit key--provided by Kristin via email on 26/2/2026.
Wrangle so we have two columns. One row per PWSID, with 2nd column indicating demand unit it pertains to. 
```{r}
du_key = read_excel("Inputs/Demand unit key/Systems_served_by_DU_pwsid.xlsx")%>%
  rename(id = 1)%>%
  mutate(id = str_trim(str_replace_all(id, "[\r\n]+", " ")),
           is_du = ifelse(startsWith(id, "CA") | grepl("San Gorgonio|Upper San Gabriel|Orange County", id), 0, 1),
         demand_unit = if_else(is_du == 1, id, NA))%>%
  fill(demand_unit, .direction = "down")%>%
  filter(is_du != 1)%>%
  rename(pwsid = 1)%>%
  select(-is_du)%>%
  separate_rows(demand_unit, sep = ";\\s*")%>%
  filter(demand_unit %in% calsim_dus$demand_unit)%>%
  mutate(demand_unit = case_when(
    #pwsid == "CA5010017" ~ "71_NU",
    pwsid == "CA4810701" ~ "TVAFB",
    pwsid == "CA1910001" ~ "ESB415",
    T ~ demand_unit))
```

GW-demand unit key
```{r}
gw_dus = read_excel("Inputs/Demand unit key/SystemswithinCalSim_groundwater.xlsx")%>%
  select(pwsid, demand_unit_final)%>%
  rename(du_gw = 2)%>%
  unique()%>%
  mutate(gw_unit = 1)
```

SABL boundary layers (in order to get the system name and location from centroid for any systems that aren't in the network dataset)
```{r}
sabl_raw = st_read("Inputs/SABL/System_Area_Boundary_Layer.shp")%>%
  rename(pwsid = 1, sabl_system_name = 4)%>%
  group_by(pwsid) %>%
  slice_max(BOUNDARY_T == "Water Service Area", n = 1)
  select(pwsid, sabl_system_name)

sabl = sabl_raw%>%
  st_centroid()%>%
  st_transform(wgs_crs)%>%
  mutate(lon_sabl = st_coordinates(geometry)[,1], 
         lat_sabl = st_coordinates(geometry)[,2])%>%
  st_drop_geometry()

sabl_filtered = sabl_raw%>%
  filter(pwsid %in% network_data$pwsid)

#st_write(sabl_filtered, "Outputs/SABL/sabl_processed.shp", delete_dsn = T)

gw_systems = sabl_raw%>%
  st_centroid()%>%
  filter(STATE_CLAS == "COMMUNITY" & !pwsid %in% network_data$pwsid)
  
st_write(gw_systems, "Outputs/GW systems/gw_systems.shp", delete_dsn = T)
```

Population served by system from SDWIS (for any systems that aren't in the network dataset)
```{r}
pop_sdwis = read_excel("Inputs/SABL/EPASDWIS_Q12025_allactiveCAPWS.xlsx", skip = 3)%>%
  rename(pwsid = 1, system_name_sdwis = 2, pop_served_sdwis = 6)%>%
  select(1, 2, 6)
```

Load in most recent version of network data (which has system lat/longs). Filter for only these relevant system variables
```{r}
network_data = read.csv("C:/Users/benji/Desktop/COEQWAL/coeqwal-benji/Data/Outputs/20260309_network_joined.csv")
```

Filter out demand units not in CalSim, and clean up/split up other demand units
```{r}
network_data_filtered = network_data%>%
  select(system_name, pwsid, calsim_demandunit, pop_served, system_lat, system_lon)%>%
  unique()%>%
  filter(!calsim_demandunit %in% c("Out of CalSim", "Out of Calsim"))%>%
  arrange(system_name)%>%
  mutate(calsim_demandunit = str_trim(str_replace_all(calsim_demandunit, "[\r\n]+", " ")))%>%
  separate_rows(calsim_demandunit, sep = ";\\s*")
```

Join network data with SW and GW demand unit keys in order to get master list
```{r}
master_key = network_data_filtered%>%
  full_join(., gw_dus, by = "pwsid")%>%
  mutate(calsim_demandunit = ifelse((calsim_demandunit == "PWSID" | is.na(calsim_demandunit)) & !is.na(du_gw), du_gw, calsim_demandunit))%>%
  full_join(., du_key%>%rename(du_sw = demand_unit), by = "pwsid")%>%
  mutate(calsim_demandunit = ifelse((calsim_demandunit == "PWSID" | is.na(calsim_demandunit)) & !is.na(du_sw), du_sw, calsim_demandunit))%>%
  unique()%>%
  arrange(pwsid)
  
gw_du_overwrite = master_key%>%
  filter(du_gw != calsim_demandunit & !is.na(du_gw))%>%
  mutate(calsim_demandunit = du_gw)
sw_du_overwrite = master_key %>%
  filter(du_sw != calsim_demandunit & !is.na(du_sw)) %>%
  mutate(calsim_demandunit = du_sw)
  
full_master_key = bind_rows(master_key, gw_du_overwrite, sw_du_overwrite) %>%
  select(system_name:system_lon)%>%
  unique()%>%
  arrange(pwsid)%>%
  mutate(gw_du = ifelse(calsim_demandunit %in% gw_dus$du_gw, 1, 0),
         sw_du = ifelse(calsim_demandunit %in% du_key$demand_unit, 1, 0),
         in_network_data = ifelse(pwsid %in% network_data$pwsid, 1, 0))

rm(gw_dus, master_key, gw_du_overwrite, sw_du_overwrite)
```

Fill in missing information from SABL and SDWIS datasets
```{r}
full_master_key = full_master_key%>%
  left_join(., sabl, by = "pwsid")%>%
  mutate(system_name = ifelse(is.na(system_name), sabl_system_name, system_name),
         system_lat = ifelse(is.na(system_lat), lat_sabl, system_lat),
         system_lon = ifelse(is.na(system_lon), lon_sabl, system_lon))%>%
  select(-c(sabl_system_name:lat_sabl))%>%
  left_join(., pop_sdwis, by = "pwsid")%>%
  mutate(system_name = ifelse(is.na(system_name), system_name_sdwis, system_name),
         pop_served = ifelse(is.na(pop_served), pop_served_sdwis, pop_served))%>%
  select(-c(system_name_sdwis:pop_served_sdwis))
```

Pick one lat/long per demand unit based on the system with the greatest population served
```{r}
lat_longs = full_master_key%>%
  group_by(calsim_demandunit)%>%
  filter(pop_served == max(pop_served, na.rm = T))%>%
  select(calsim_demandunit, pwsid, system_name, pop_served:sw_du)%>%
  st_as_sf(., coords = c("system_lon", "system_lat"), crs = wgs_crs)%>%
  st_transform(ca_crs)%>%
  st_jitter(., amount = 500)%>%
  st_transform(wgs_crs)%>%
  mutate(lon = st_coordinates(geometry)[,1], 
         lat = st_coordinates(geometry)[,2],
         dw_du = ifelse(calsim_demandunit %in% calsim_dus$demand_unit, 1, 0))%>%
  st_drop_geometry()
```

Export
```{r}
write.csv(full_master_key, "Outputs/Demand unit to PWSID keys/20260324_demand_unit_pwsid_master_list.csv", row.names = F)
write.csv(lat_longs, "Outputs/Demand unit to PWSID keys/20260324_demand_unit_lat_longs.csv", row.names = F)
```