

# Set up

Import packages
```{r}
suppressPackageStartupMessages({
  suppressWarnings({
    library(tidyverse)
    library(readxl)
    library(here)
    library(sf)
  })})
```

Set settings and universal variables
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here())

options(scipen = 999)
```

# Import network data
```{r}
network_df = read_excel("C:/Users/benji/Downloads/Network data with updated calsim demand assignments - April 3rd.xlsx")%>%
  mutate(average_source_usage = average_source_usage / 100)
```

# Process data

Create new rows to fill in the remainder (to hit 100% volume) for each system
```{r}
# Combine with original data
network_df = bind_rows(
  network_df, 
  network_df %>%
    group_by(pwsid) %>%
    reframe(
      source_name = "gw", 
      source_type = "gw",
      calsim_demandunit = "gw",
      average_source_usage = 1 - sum(average_source_usage))%>%
    filter(average_source_usage > 0))
```

# Demand unit tracing

Create a dataframe of 'end nodes' that have a demand unit (and therefre don't need to be traced back further)
```{r}
end_nodes <- network_df %>%
  filter(calsim_demandunit != "PWSID") %>%
  select(pwsid, traced_du = calsim_demandunit, average_source_usage)%>%
  group_by(pwsid, traced_du)%>%
  reframe(average_source_usage = sum(average_source_usage))
```

Recursive function to trace back when any connections have their demand unit listed as 'PWSID'
```{r}
# Step 1: Build PWSID → DU mappings (terminal nodes)
du_mappings <- network_df %>%
  filter(calsim_demandunit != "PWSID") %>%
  select(pwsid, source = calsim_demandunit, volume = average_source_usage)

# Step 2: Build PWSID → PWSID mappings (intermediate nodes)
pwsid_mappings <- network_df %>%
  filter(calsim_demandunit == "PWSID") %>%
  select(pwsid, source = source_name, volume = average_source_usage)

# Step 3: Combine and calculate fractions of volume
complete_map <- bind_rows(du_mappings, pwsid_mappings) %>%
  group_by(pwsid, source) %>%
  summarise(volume = sum(volume), .groups = "drop") %>%
  group_by(pwsid) %>%
  mutate(fraction = volume / sum(volume)) %>%
  ungroup()

# Step 4: Create lookup list
complete_list <- complete_map %>%
  group_by(pwsid) %>%
  group_split() %>%
  setNames(map_chr(., ~ first(.x$pwsid))) %>%
  map(~ select(.x, source, fraction))

# Step 5: Recursive trace function
trace_to_dus <- function(source_name, volume = 1, complete_list, visited = c()) {
  # Prevent infinite loops from circular dependencies
  if (source_name %in% visited) {
    warning(paste("Circular reference detected:", source_name))
    return(tibble(traced_du = source_name, adjusted_volume = volume))}
  
  # If this source is a PWSID that can be decomposed
  if (source_name %in% names(complete_list)) {
    components <- complete_list[[source_name]]
    
    results <- map2_dfr(
      components$source,
      components$fraction,
      ~ trace_to_dus(.x, volume * .y, complete_list, c(visited, source_name)))
    return(results)} 
  else {
    # End node, so stop recursion and record that endnode's info
    return(tibble(traced_du = source_name, adjusted_volume = volume))}}

# Step 6: Apply to all purchases
purchases <- network_df %>%
  filter(calsim_demandunit == "PWSID") %>%
  select(pwsid, source_name, average_source_usage)

traced_results <- purchases %>%
  rowwise() %>%
  mutate(
    traced = list(trace_to_dus(source_name, average_source_usage, complete_list))) %>%
  unnest(traced) %>%
  select(pwsid, traced_du, adjusted_volume) %>%
  group_by(pwsid, traced_du) %>%
  summarise(average_source_usage = sum(adjusted_volume), .groups = "drop")

# Step 7: combine traced results with end nodes to get 'master sheet' of each system's complete volume distribution across demand units
final = rbind(traced_results, end_nodes)%>%
  mutate(average_source_usage = average_source_usage *100)%>%
  group_by(pwsid, traced_du)%>%
  reframe(average_source_usage = sum(average_source_usage))
```

# Diagnostics

Do some manual tracing to make sure everything was calculated correctly
```{r}
final%>%
  filter(pwsid == "CA0110003")
```

```{r}
network_df%>%
  mutate(average_source_usage = average_source_usage * 100)%>%
  filter(pwsid %in% c("CA0110003", "CA0110010"))%>%
  select(pwsid, source_name, average_source_usage, calsim_demandunit)
```

# Export
```{r}
write.csv(final, "Outputs/Static demand/20260406_traced_demand_units.csv", row.names = F)
```


